Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-11-09 doc/paper.tex
add footnote
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>
update paper?
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large. Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed. In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem. The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems. Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks. In recent years, LLMs have been successfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to. We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other.

@@

The agent is trained as in the baseline. The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep. In addition to the state observation, the LLM is also asked to respond with a list of recommended actions for the agent to perform. The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with. As it does so, its chosen action is encoded to natural text. Then, the LLM is used as a tool for semantic similarity comparison, comparing its own formerly proposed action to the action performed by the agent. Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment. Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropogation after an entire episode is completed. Thus, this method functions more as an advisor, influencing the policy in a more long term capacity than the coaching of the policy influencer. Throughout the paper, we will refer to this as LLM reward shaping.

@@

In both these cases, we need two language models; one to generate the suggestions and one to generate semantic similarity values.

For text generation, we have decided utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023. Specifically, we chose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}. In the experiments presented below, we use the smallest version of Llama 2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call an API to a model hosted by someone else. The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.

Some research papers propose fine-tuning of LLMs as means of leveraging the capabilites of the language models for decision-making purposes even further \cite{grounding}. Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning. From the perspective of reinforcement learning, this is the appropriate course of action, as the LLM should not possess any previous knowledge about the task at hand, just like the agent should not. In addtion, the act of fine-tuning LLM is a rather time-comsuming process, requiring the generation and structuring of a new dataset, and also vast hardware resources.

@@

For generating semantic similarity values, we use Sentence-Bert \cite{bert} for generating encoding vectors, and perform a cosine similarity calculation of the two sentence encodings. This gives a value between $-1$ and $1$, where two sentences with the exact same encoding have a cosine similarity of $1$.

@@

We select three different Minigrid environments with increasing complexity; Empty, DoorKey and UnlockPickup. For each environment we test our four different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy influencing. All methods use the same hyperparameters for the PPO, which can be found in the appendix. For each of these methods, we run PPO for 2000 (200 for Empty) episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one. We do this eight times per method, and compare the average results per episode against each other.

@@

In this simple environment, we see that both LLM reward shaping, LLM policy influencing and A2C2 outperform the baseline model. Averaging over all episodes for all eight runs, we find that reward shaping increases the mean reward received by $10\%$ over the baseline, with a lower standard deviation in rewards received per run of $0.12$. Policy influencing scores even higher at $32\%$ over the baseline, with an even lower standard deviation of $0.09$. Finally, A2C2 performs the best of the four with a $37\%$ increase in average reward over the baseline model, and the lowest standard deviation of $0.04$

@@

\begin{figure}[h]
\centerline{\includegraphics[width=3.25in]{figure/emptyresults.png}}
\caption{Reward per episode for the four different methods in the Empty environment}
\label{doorkeyresults}
\end{figure}

\begin{table}[h]
\caption{All results on Empty environment}
\begin{center}
\label{emptytable}
\begin{tabular}{c | c c c}
Method & Mean reward & Std reward & Increase over baseline \\
\hline
Reward shaping & 0.76 & 0.12 & 10\% \\
Policy influencing & 0.91 & 0.09 & 32\% \\
Both & 0.95 & 0.04 & 37\% \\
Baseline & 0.69 & 0.28 & 0\% \\
\end{tabular}
\end{center}
\end{table}

@@

As we can see from Figure \ref{doorkeyresults}, we see impressive results from our LLM integration in this environment as well, especially from reward shaping and A2C2. For the mean and standard deviation of all methods, see table \ref{doorkeytable}.

\begin{table}[h]
\caption{All results on DoorKey environment}
\begin{center}
\label{doorkeytable}
\begin{tabular}{c | c c c}
Method & Mean reward & Std reward & Increase over baseline \\
\hline
Reward shaping & 0.48 & 0.28 & 24\% \\
Policy influencing & 0.36 & 0.31 & -6\% \\
Both & 0.53 & 0.25 & 38\% \\
Baseline & 0.38 & 0.31 & 0\% \\
\end{tabular}
\end{center}
\end{table}

@@

The Unlock Pickup environment is an even more complex environment than DoorKey, requiring a much longer string of actions to give out a reward. As may be anticipated, this added complexity makes the average reward much lower, and can see from Fig. \ref{unlockpickupresults} that all methods struggle (keep in mind the the y-axis now only goes to $0.15$, as opposed to the maximum reward possible of $1$, which the other graphs use).

@@

\begin{table}[h]
\caption{All results on UnlockPickup environment}
\begin{center}
\label{unlockpickuptable}
\begin{tabular}{c | c c c}
Method & Mean reward & Std reward & Increase over baseline \\
\hline
Reward shaping & 0.037 & 0.082 & 275\% \\
Policy influencing & 0.014 & 0.034 & 43\% \\
Both & 0.031 & 0.072 & 222\% \\
Baseline & 0.0097 & 0.025 & 0\% \\
\end{tabular}
\end{center}
\end{table}

Looking at figure \ref{unlockpickupresults}, we can see that policy influencing also struggles in this environment. From table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in the reward shaping method. Policy influencing received a mean average score of $0.014$. In this complicated environment, it may that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up. As explained earlier, this instantly loses the game by removing the box from the game. Thus, while the results are better than the baseline, they are still very poor and we do not consider policy influencing to be particularly beneficial in this environment.

@@

Having looked at the performance of our different models, we can see that the suggestions from the LLM have positive effects in many cases, while also sometimes being detrimental. We now inspect some sample responses from the LLM, to gain deeper understanding of how the LLM is effecting the agent.

@@

\begin{table}[h]
\caption{Sample responses from LLM in Empty environment.}
\begin{center}
\label{emptyresponses}
\begin{tabularx}{3.25in}{X | X}
Prompt & Response \\
\hline
My goal is: get to the green goal square. I see a green goal left and forward & move forward. \\
My goal is: get to the green goal square. I see a green goal left. & move left. \\
My goal is: get to the green goal square. I see a green goal directly in front. & move forward.
\end{tabularx}
\end{center}
\end{table}

@@

From table \ref{doorkeyresponses} we can see that some suggestions are good, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present. We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the players inventory. These poor responses can explain why policy influencing can have a detrimental impact in this environment, as these actions will always be sampled slightly more often.

@@

On our final environment, we can see (Table \ref{unlockpickupresponses}) beneficial responses similar to the ones given in DoorKey. Here, we also see some absurd responses, which show that the LLM has completely misunderstood the prompt and few-shot examples. These responses are not particularly common, but they do occur sometimes and reveal some of the problems with running a relatively small LLM locally, as opposed to a larger model such as GPT4.

\begin{table}[h]
\caption{Sample responses from LLM in UnlockPickup environment. Good responses above line and suboptimal responses below.}
\begin{center}
\label{unlockpickupresponses}
\begin{tabularx}{3.25in}{X | X}
Prompt & Response \\
\hline
My goal is: pick up the blue box. I see a locked red door LEFT and FORWARD, see a red key directly in front. & pick up red key. \\
\hline
My goal is: pick up the grey box. I have a grey key in my inventory. & Congratulations, you have successfully reached your goal! You find a small, mysterious box inside the grey box. It's made of a strange, otherworldly material that seems to glow with an ethereal light. As you examine it closer, you can feel a faint hum of energy emanating from it. \\
My goal is: pick up the green box. I have a purple key in my inventory. & Great! With the purple key in your inventory, you can now unlock the green box. You approach the green box and use the purple key to unlock it. Inside, you find a shiny coin!
\end{tabularx}
\end{center}
\end{table}

@@

\begin{table}[h]
\caption{PPO hyperparameters}
\begin{center}
\label{hyperparams_transposed}
\begin{tabular}{c | c}
Parameter & Value \\
\hline
Gamma & 0.99 \\
Steps/episode & 1024 \\
Epochs/episode & 10 \\
Batches/epoch & 8 \\
Clip & 0.2 \\
Value loss coeff. & 0.3 \\
Entropy coeff. & 0 \\
Learning rate & 3.0e-4 \\
\end{tabular}
\end{center}
\end{table}

>>>>

Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large. Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed. In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem. The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems. Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks. In recent years, LLMs have been successfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to. We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other. \footnote{Code available at \href{https://github.com/jonatan-hanssen/in5490-project}{GitHub}}

@@

The agent is trained as in the baseline. The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep. After being given the state observation, the LLM is asked to respond with a list of recommended actions for the agent to perform. The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with. As it does so, its chosen action is encoded to natural text. Then, another LLM is used for semantic similarity comparison, comparing the proposed action to the action performed by the agent. Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment. Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropogation after an entire episode is completed. Thus, this method functions more as an advisor, influencing the policy in a more long term capacity than the coaching of the policy influencer. Throughout the paper, we will refer to this as LLM reward shaping.

@@

In all these cases, we need two language models; one to generate the suggestions and one to generate semantic similarity values.

For text generation, we have decided to utilize an LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023. Specifically, we chose Llama 2, which has been trained using 40\% more data than its predecessor Llama 1, and has double the context length\cite{llama}. In the experiments presented below, we use the smallest version of Llama 2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect and cost of having to call an API to a model hosted by someone else. The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.

Some research papers propose fine-tuning of LLMs as a means of leveraging the capabilites of the language models for decision-making purposes even further \cite{grounding}. Our plan of action, on the other hand, aims at evaluating the efficacy of the LLM without any previous pre-tuning. From the perspective of reinforcement learning, this is the appropriate course of action, as the LLM should not possess any previous knowledge about the task at hand, just like the agent should not. In addtion, the act of fine-tuning LLM is a rather time-comsuming process, requiring the generation and structuring of a new dataset, and also vast hardware resources.

@@

For generating semantic similarity values, we follow \cite{ellm} and use Sentence-Bert \cite{bert} for generating encoding vectors, and perform a cosine similarity calculation of the two sentence encodings. This gives a value between $-1$ and $1$, where two sentences with the exact same encoding have a cosine similarity of $1$.

@@

We select three different Minigrid environments with increasing complexity; Empty, DoorKey and UnlockPickup. For each environment we test our four different models; PPO with no LLM integration, PPO with LLM reward shaping, PPO with LLM policy influencing and PPO with both reward shaping and policy influencing (A2C2). All methods use the same hyperparameters for the PPO, which can be found in the appendix. For each of these methods, we run PPO for 2000 (200 for Empty) episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one. We do this eight times per method, and compare the average results per episode against each other.

@@

In this simple environment, we see from Figure \ref{emptyresults} that both LLM reward shaping, LLM policy influencing and A2C2 outperform the baseline model. Averaging over all episodes for all eight runs, we find that reward shaping increases the mean reward received by $10\%$ over the baseline, with a lower standard deviation in rewards received per run of $0.12$. Policy influencing scores even higher at $32\%$ over the baseline, with an even lower standard deviation of $0.09$. Finally, A2C2 performs the best of the four with a $37\%$ increase in average reward over the baseline model, and the lowest standard deviation of $0.04$. See Table \ref{emptytable}.

@@

\begin{figure}[h]
\centerline{\includegraphics[width=3.25in]{figure/emptyresults.png}}
\caption{Reward per episode for the four different methods in the Empty environment}
\label{emptyresults}
\end{figure}

\begin{table}[h]
\caption{All results on Empty environment}
\begin{center}
\label{emptytable}
\begin{tabular}{c | c c c}
Method & Mean reward & Std reward & Increase over baseline \\
\hline
Reward shaping & 0.76 & 0.12 & 10\% \\
Policy influencing & 0.91 & 0.09 & 32\% \\
Both & 0.95 & 0.04 & 37\% \\
Baseline & 0.69 & 0.28 & - \\
\end{tabular}
\end{center}
\end{table}

@@

As we can see from Figure \ref{doorkeyresults}, LLM integration gives impressive results in this environment as well, especially from reward shaping and A2C2. For the mean and standard deviation of all methods, see table \ref{doorkeytable}.

\begin{table}[h]
\caption{All results on DoorKey environment}
\begin{center}
\label{doorkeytable}
\begin{tabular}{c | c c c}
Method & Mean reward & Std reward & Increase over baseline \\
\hline
Reward shaping & 0.48 & 0.28 & 24\% \\
Policy influencing & 0.36 & 0.31 & -6\% \\
Both & 0.53 & 0.25 & 38\% \\
Baseline & 0.38 & 0.31 & - \\
\end{tabular}
\end{center}
\end{table}

@@

The Unlock Pickup environment is an even more complex environment than DoorKey, requiring a much longer string of actions to give out a reward. As may be anticipated, this added complexity makes the average reward much lower, and we can see from Figure \ref{unlockpickupresults} that all methods struggle (keep in mind the the y-axis now only goes to $0.15$, as opposed to the maximum reward possible of $1$, which the other graphs use).

@@

\begin{table}[h]
\caption{All results on UnlockPickup environment}
\begin{center}
\label{unlockpickuptable}
\begin{tabular}{c | c c c}
Method & Mean reward & Std reward & Increase over baseline \\
\hline
Reward shaping & 0.037 & 0.082 & 275\% \\
Policy influencing & 0.014 & 0.034 & 43\% \\
Both & 0.031 & 0.072 & 222\% \\
Baseline & 0.0097 & 0.025 & - \\
\end{tabular}
\end{center}
\end{table}

Looking at figure \ref{unlockpickupresults}, we can see that policy influencing also struggles in this environment. From table \ref{unlockpickuptable}, we see a $43\%$ increase, although this is not very impressive given the low score of the baseline, and it is difficult to see an upward trend in the amount of reward received, which can be clearly seen in both reward shaping and A2C2. Policy influencing received a mean score of $0.014$. In this complicated environment, it may be that suboptimal responses from the LLM have high chance to ruin the agent's chance to complete the goal, for example by suggesting to open the box instead of picking it up. As explained earlier, this instantly loses the game by removing the box from the game. Thus, while the results are better than the baseline, they are still very poor and we do not consider policy influencing to be particularly beneficial in this environment.

@@

Having looked at the performance of our different models, we can see that the suggestions from the LLM have positive effects in many cases, while also sometimes being detrimental. We now inspect some sample responses from the LLM, to gain deeper understanding of how the LLM is affecting the agent.

@@

\begin{table}[h]
\caption{Sample responses from LLM in Empty environment.}
\begin{center}
\label{emptyresponses}
\begin{tabularx}{3.25in}{X | X}
Prompt & Response \\
\hline
My goal is: get to the green goal square. I see nothing interesting. & move forward. \\
My goal is: get to the green goal square. I see a green goal left and forward. & move forward. \\
My goal is: get to the green goal square. I see a green goal left. & move left. \\
My goal is: get to the green goal square. I see a green goal directly in front. & move forward.
\end{tabularx}
\end{center}
\end{table}

@@

From table \ref{doorkeyresponses} we can see that some suggestions are good, such as suggestion to pick up a key when it is directly in front, and moving towards the goal although other observations are also present. We also see some suboptimal responses, such as suggesting to pick up a key when it is not directly in front and suggesting to pick up a key when it is already in the player's inventory. These poor responses can explain why policy influencing can have a detrimental impact in this environment, as these actions will always be sampled slightly more often.

@@

On our final environment, we can see (Table \ref{unlockpickupresponses}) beneficial responses similar to the ones given in DoorKey. Here, we also see some absurd responses, which show that the LLM has completely misunderstood the prompt and few-shot examples. These responses are not particularly common, but they do occur sometimes and reveal some of the problems with running a relatively small LLM locally, as opposed to a larger model such as GPT4. These problems have been attempted remedied through prompt engineering (see the Appendix).

\begin{table}[h]
\caption{Sample responses from LLM in UnlockPickup environment. Good responses above line and suboptimal responses below.}
\begin{center}
\label{unlockpickupresponses}
\begin{tabularx}{3.25in}{X | X}
Prompt & Response \\
\hline
My goal is: pick up the blue box. I see a locked red door left and forward, see a red key directly in front. & pick up red key. \\
\hline
My goal is: pick up the grey box. I have a grey key in my inventory. & Congratulations, you have successfully reached your goal! You find a small, mysterious box inside the grey box. It's made of a strange, otherworldly material that seems to glow with an ethereal light. As you examine it closer, you can feel a faint hum of energy emanating from it. \\
My goal is: pick up the green box. I have a purple key in my inventory. & Great! With the purple key in your inventory, you can now unlock the green box. You approach the green box and use the purple key to unlock it. Inside, you find a shiny coin!
\end{tabularx}
\end{center}
\end{table}

@@

\subsection{Prompt given to Llama2}

Below is the system prompt sent to Llama2, which is prepended before any state observation:

{\it You are a helpful assistant giving advice to someone playing a videogame. You will recieve the current goal of the game and a list of observations about the objects in the environment and their positions relative to the player at one single timestep. You shall give a list of suggested actions that the player will take to reach their goal during this single timestep. The available actions to the player are: move forward, turn left and right, pick up items, drop items and use items. You MUST ALWAYS RESPOND with one or more of these actions. Your response MUST ALWAYS follow the template of the given example responses. NEVER ADD anything not in the example responses. You will NEVER ASSUME that other possible actions exist. You will NEVER ASSUME that the game is finished. You WILL NEVER make any other assumptions about the environment, only use the information given to you. If none of the observations are relevant to solving the task, respond that the player should turn around and explore more. Separate each suggestion with a new line. Be concise. ONLY FOLLOW THESE INSTRUCTIONS. NEVER RESPOND IN ANY OTHER WAY. OTHERWISE, YOU WILL BE SEVERELY PUNISHED.}

\subsection{Figures and tables}

\begin{table}[h]
\caption{PPO hyperparameters}
\begin{center}
\label{hyperparams_transposed}
\begin{tabular}{c | c}
Parameter & Value \\
\hline
Gamma & 0.99 \\
Steps per episode & 1024 \\
Epochs per episode & 10 \\
Batches per epoch & 8 \\
PPO Clipping value & 0.2 \\
Value loss coefficient & 0.3 \\
Entropy coefficient. & 0 \\
Learning rate & 3.0e-4 \\
\end{tabular}
\end{center}
\end{table}
